\bf Model Visualization Techniques for a Social Network Model

### 2017-04-17 paper.Rnw
light editing to intro

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }

Models for dynamic social networks are extremely important because of how realistic they are. Social networks do not form spontaneously; they evolve over time. Ties are added and removed, and new actors can join the network. Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.

\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) of \citet{saompaper}. These models are uniquely different from other social network models because they incorporate network statistics and actor statistics into the model, whereas most other models just use the network statistics to determine change. Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change. In the "real-world" we would expect that people with common interests would be more likely to form relationships, and that intuition is no longer ignored in modeling with SAOMs.

\par The software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models. This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works. To better understand the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization from \citet{modelvis}. With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic. By bringing some light to the underlying methods and structures "behind the scenes" of SAOMs and RSiena, we will help other researchers working with the models better understand and better use these models.

\par The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model. Again, in these steps are largely hidden from the researcher computing them with software. We hope that by bringing these background elements and others into the foreground with static and dynamic visualizations, stochastic actor-oriented models will be better understood and more accessible to social network analysts.

>>>>

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}

Models for dynamic social networks are extremely important because of how realistic they are. Social networks do not form spontaneously; they evolve over time. Ties are added and removed, and new actors join the network. Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.

\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they incorporate network {\emph and} actor statistics into the model, when other models only rely on  the network statistics to determine change. Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change. In the ``real" world we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporates this intuition  in the modeling process.

\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis.%, and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
However, unlike other models, SAOMs are not very well understood. To get better insight into the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs and RSiena, we  help researchers working with the models better understand and better use these models.

\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model. Again, these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.
